---
title: "Data VS 2"
author: "Mai Ngo"
date: "2024-04-19"
output: pdf_document
---

## Note: I am approaching this problem considering data analytic as well. Considering time series data, I will plot moving average at both month and daily level. 

```{r}
setwd("C:/Users/maimu/OneDrive/Documents/DePaul/DSC 465/2nd")
```

#Libraries.
```{r message=FALSE}
library(dplyr) 
library(ggplot2)
library(lubridate)
```

Read data sets. Get basic information about the Portland Water Level data set.
```{r, results='asis'}
waterData <- read.csv("PortlandWaterLevel2003.csv", header=TRUE, sep=",")
#Drop columns no need for graph.
waterData <- waterData[, -which(names(waterData) %in% c('Station','Sigma','I','L'))]
waterData[waterData == ""] <- NA

print("Portland Water Level Data:")
print(paste("Missing values count: ", sum(is.na(waterData))))
print(paste("Row:", dim(waterData)[1], "Column:", dim(waterData)[2]))
```

# Q3a:

## 1st approach: Using Tableau formula - Average of previous 2-month sum, next 0 including current observations.

```{r}
#Get sum of Water Level for each Month.
waterData$Date <- mdy(waterData$Date) 
waterData$Month <- month(waterData$Date, label = TRUE)
waterData_monthlySum <- waterData %>% group_by(Month) %>% summarise(sumWL = sum(WL))

#Calculate moving average manually.
waterData_monthlySum$Moving_Avg <- NA 
waterData_monthlySum$Moving_Avg[1] <- waterData_monthlySum$sumWL[1]
#Second obs: average of current and first.
waterData_monthlySum$Moving_Avg[2] <- mean(waterData_monthlySum$sumWL[1:2])

#Third obs onward: average of current obs and previous two.
for (m in 3:nrow(waterData_monthlySum))
{waterData_monthlySum$Moving_Avg[m] <-mean(waterData_monthlySum$sumWL[(m-2):m])}
waterData_monthlySum
```

##Line plot.
```{r, message=FALSE}
overallAvg <- mean(waterData_monthlySum$Moving_Avg)

ggplot(data = waterData_monthlySum, aes(x = Month, y = Moving_Avg, group =1)) +
  geom_line(color='red') +    
  geom_point(color='red') +
  geom_smooth(color='darkgreen', se=F) +
  scale_y_continuous(limits = c(0, NA)) +
  geom_text(aes(label = round(Moving_Avg, 2)), vjust=3, size=3, color="black") +  
  geom_hline(yintercept = overallAvg, linetype = "dashed", color="blue") + 
  theme_minimal()+
  labs(title = "1st Approach - Tableau formula: \nMonthly Moving Average of Water Level in Portland (2003).",
       x = "Month", y = "Water Level Moving Average")
```

## 2nd: Monthly Simple Moving Average. My own formula which discussed during class.

1. Calculate the moving average for each observation, previous two + current observations. 
--> Discussed in-class and got approved.
2. Average of each month in 2 ways: with(2a) and without(2b) current observation.
--> This is because the month of December only have day observation of December 1st. Including this date's observations will result a drop in water level throughout the year.

### 1.Calculate the moving average for each observation, previous two + current observations. 
```{r}
waterData$Datetime <- ymd_hm(paste(waterData$Date, waterData$Time))

#Same logic as first approach. Only difference is instead of by month, we do by hour.
waterData$Moving_Avg <- NA 
for (m in 7:nrow(waterData)) {waterData$Moving_Avg[m] <- mean(waterData$WL[(m-6):m])}
waterData$Moving_Avg[1] <- waterData$WL[1]
waterData$Moving_Avg[2] <- mean(waterData$WL[1:2])
waterData$Moving_Avg[3] <- mean(waterData$WL[1:3])
waterData$Moving_Avg[4] <- mean(waterData$WL[1:4])
waterData$Moving_Avg[5] <- mean(waterData$WL[1:5])
waterData$Moving_Avg[6] <- mean(waterData$WL[1:6])
```

### 2a. Average of each month INCLUDING current observation.
```{r}
#Using sum following Tableau formula.
waterData_incCurrent <- waterData %>% group_by(Month) %>% 
                     summarise(Month_Avg = sum(Moving_Avg), .groups = "drop")

#Taking average with 2 previous months.
waterData_incCurrent$Moving_Avg_Prev2 <- NA 
waterData_incCurrent$Moving_Avg_Prev2[1] <- waterData_incCurrent$Month_Avg[1]
#Second obs: average of current and first.
waterData_incCurrent$Moving_Avg_Prev2[2] <- mean(waterData_incCurrent$Month_Avg[1:2])
#Third obs onward: average of current obs and previous two.
for (m in 3:nrow(waterData_incCurrent)) 
{waterData_incCurrent$Moving_Avg_Prev2[m] <- 
    mean(waterData_incCurrent$Month_Avg[(m-2):m])}

#Taking average with 3 previous months.
waterData_incCurrent$Moving_Avg_Prev3 <- waterData_incCurrent$Moving_Avg_Prev2
#Third obs.
waterData_incCurrent$Moving_Avg_Prev3[3] <- mean(waterData_incCurrent$Month_Avg[1:3])
#Third obs onward: average of current obs and previous two.
for (m in 4:nrow(waterData_incCurrent)) 
{waterData_incCurrent$Moving_Avg_Prev3[m] <- 
    mean(waterData_incCurrent$Month_Avg[(m-3):m])}

#Taking average with all previous months.
waterData_incCurrent$Moving_Avg_prevAll<-cummean(waterData_incCurrent$Month_Avg)
print(waterData_incCurrent)
```

##Line plot including current month observation for moving average.

Comment: 

* The graph represents Monthly Moving Average of Water Level in Portland (2003) including current month data in moving average calculation.
* The black line represent average water level of every month, considering water level were re-calculated as moving average with previous two hours. 
* Three smoothing lines represents different moving average methods. Overall, moving average includes all previous months yields the most smoothed line. While moving average consider the previous 2-3 months shows a light increase in September.
* The graph holds expectation that there is a light drop in December due to insufficient data of the month.

```{r, warning=FALSE, message=FALSE}
overallAvg <- mean(waterData_incCurrent$Month_Avg)
ggplot(data = waterData_incCurrent
       , aes(x = Month, y = Month_Avg, group = 1)) +
  geom_line(color = 'black',size = 0.5) +    
  geom_hline(yintercept = overallAvg, linetype = "dashed", color = "blue") + 
  geom_smooth(aes(y = Moving_Avg_Prev2, colour = 'Previous 2'), se = FALSE) +
  geom_smooth(aes(y = Moving_Avg_Prev3, colour = 'Previous 3'), se = FALSE) +
  geom_smooth(aes(y = Moving_Avg_prevAll, colour = 'All previous'), se = FALSE) +
  scale_y_continuous(limits = c(0, NA)) +
  scale_colour_manual(values=c("darkgreen", "darkorange", "purple"),
                      name = "Moving average including current month") +
  theme_minimal() +
  theme(legend.position="bottom") +
  labs(title = "Monthly Moving Average of Water Level in Portland (2003)",
       x = "Month",
       y = "Water Level Moving Average")
```

### 2b. Average of each month EXCLUDING current observation.
```{r}
waterData_excCurrent <- waterData %>% group_by(Month) %>% 
                     summarise(Month_Avg = sum(Moving_Avg), .groups = "drop")

#Taking average with 2 previous months.
waterData_excCurrent$Moving_Avg_Prev2 <- NA 
waterData_excCurrent$Moving_Avg_Prev2[1] <- waterData_incCurrent$Month_Avg[1]
waterData_excCurrent$Moving_Avg_Prev2[2] <- waterData_incCurrent$Month_Avg[2]
#Third obs onward: average of previous two.
for (m in 3:nrow(waterData_excCurrent)) 
{waterData_excCurrent$Moving_Avg_Prev2[m] <- 
    mean(waterData_excCurrent$Month_Avg[(m-2):(m-1)])}

#Taking average with 3 previous months.
waterData_excCurrent$Moving_Avg_Prev3 <- waterData_excCurrent$Moving_Avg_Prev2
#Third obs.
waterData_excCurrent$Moving_Avg_Prev3[3] <- mean(waterData_excCurrent$Month_Avg[1:2])
#Third obs onward: average of current obs and previous two.
for (m in 4:nrow(waterData_excCurrent)) 
{waterData_excCurrent$Moving_Avg_Prev3[m] <- 
    mean(waterData_excCurrent$Month_Avg[(m-3):(m-1)])}

#Taking average with all previous months.
prev_monthSum <- cumsum(waterData_excCurrent$Month_Avg) - waterData_excCurrent$Month_Avg
prev_monthN <- 1:nrow(waterData_excCurrent) - 1
#Cumulative mean excluding current month.
waterData_excCurrent$Moving_Avg_prevAll <- prev_monthSum / prev_monthN
waterData_excCurrent$Moving_Avg_prevAll[1] <- waterData_incCurrent$Month_Avg[1]

print(waterData_excCurrent)
```

##Line plot excluding current month observation for moving average.

Comment: 

* The graph represents Monthly Moving Average of Water Level in Portland (2003) excluding current month data in moving average calculation.
* The black line represent average water level of every month, considering water level were re-calculated as moving average with previous two hours. 
* Three smoothing lines represents different moving average methods. Overall, moving average includes all previous months yields the most smoothed line with expectation of high-water level in December. While moving average consider the previous 2-3 months shows a light increase in May and light decrease in August.
* Compared to above graph, this graph has more of time-series prediction characteristic due to not including December data. Show trend that December could have high water level, while above graph displays the insufficiency data points of December hence, the drop.

```{r warning=FALSE, message=FALSE}
overallAvg <- mean(waterData_excCurrent$Month_Avg)
ggplot(data = waterData_excCurrent
       , aes(x = Month, y = Month_Avg, group =1)) +
  geom_line(color = 'black',size = 0.5) +    
  geom_hline(yintercept = overallAvg, linetype = "dashed", color = "blue") + 
  geom_smooth(aes(y = Moving_Avg_Prev2, colour = 'Previous 2'), se = FALSE) +
  geom_smooth(aes(y = Moving_Avg_Prev3, colour = 'Previous 3'), se = FALSE) +
  geom_smooth(aes(y = Moving_Avg_prevAll, colour = 'All previous'), se = FALSE) +
  scale_y_continuous(limits = c(0, NA)) +
  scale_colour_manual(values=c("darkgreen", "darkorange", "purple"),
                      name = "Moving average excluding current month") +
  theme_minimal() +
  theme(legend.position="bottom") +
  labs(title = "Monthly Moving Average of Water Level in Portland (2003)",
       x = "Month",
       y = "Water Level Moving Average")
```

## 3rd: Daily Simple Moving Average. Look further at daily level -- suggested by professor.

### Average of each day.
Note: Moving average for two and three previous days yield the same pattern.
```{r}
waterData_dayAvg <- waterData %>% group_by(Date) %>% summarise(Day_Avg = mean(WL))

#Taking average with 2 previous months.
waterData_dayAvg$Moving_Avg_Prev2 <- NA 
waterData_dayAvg$Moving_Avg_Prev2[1] <- waterData_dayAvg$Day_Avg[1]
#Second obs: average of current and first.
waterData_dayAvg$Moving_Avg_Prev2[2] <- mean(waterData_dayAvg$Day_Avg[1:2])
#Third obs onward: average of current obs and previous two.
for (m in 3:nrow(waterData_dayAvg)) 
{waterData_dayAvg$Moving_Avg_Prev2[m] <- 
    mean(waterData_dayAvg$Day_Avg[(m-2):m])}

#Taking average with all previous months.
waterData_dayAvg$Moving_Avg_prevAll<-cummean(waterData_dayAvg$Day_Avg)
head(waterData_dayAvg,5)
```

##Line plot including current date observation for moving average.

Comment: 

* The graph represents Daily Moving Average of Water Level in Portland (2003) including current day data in moving average calculation.
* The black line represent average water level of every day. We can see there are fluctuations between the days with peak at mid-March, and lowest water level at mid-May.
* Two smoothing lines represents different moving average methods. Overall, moving average includes all previous months yields the most smoothed line with showcase trend in decreasing water level throughout the year. While moving average consider the previous 2 days shows a light  decrease mid-June and slowly increase for the rest of the year.
* Compared to above monthly graphs, this graph gives more detail on water cycle at daily level. Overall pattern is that water level slowly decreasing over the year.

```{r warning=FALSE, message=FALSE}
overallAvg <- mean(waterData_dayAvg$Day_Avg)
ggplot(data = waterData_dayAvg
       , aes(x = Date, y = Day_Avg, group = 1)) +
  geom_line(color = 'black',size = 0.5) +    
  geom_hline(yintercept = overallAvg, linetype = "dashed", color = "blue") + 
  geom_smooth(aes(y = Moving_Avg_Prev2, colour = 'Previous 2'), se = FALSE) +
  geom_smooth(aes(y = Moving_Avg_prevAll, colour = 'All previous'), se = FALSE) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b %e") +
  scale_y_continuous(limits = c(0, NA)) +
  scale_colour_manual(values=c("darkgreen", "darkorange"),
                      name = "Moving average including current day") +
  theme_minimal() +
  theme(legend.position="bottom") +
  labs(title = "Daily Moving Average of Water Level in Portland (2003)",
       x = "Day",
       y = "Water Level Moving Average")
```
